---
output:
  html_document:
    pandoc_args: './common.yaml'
params:
  config.args: "./config.yaml"
---

# Part 5: Marker genes

```{r setup_opt, warning=FALSE, error=FALSE, message=FALSE, echo = F, eval=F}
options(width = 450)
```

## Load libraries
```{r load_libraries, warning=FALSE, error=FALSE, message=FALSE}
library(Seurat)
library(ggplot2)
library(dplyr)
library(future)
library(kableExtra)
library(cowplot)
library(yaml)
```

```{r prepare_inline_variables, echo = FALSE}
# get the datetime and set it to variable
todays.date <- Sys.Date()
formatted.date <- format(todays.date, format="%B %d %Y")

config.args <- read_yaml(params$config.args)
```


```{r user_vals, warning=FALSE, error=FALSE, message=FALSE, echo = FALSE}
r.obj.loc <- config.args$data$`rds-file-path`

resolution.selection <- config.args$analysis$part5$resolution_selection

run_tests <- config.args$analysis$run_tests
cell.cycle.string <- config.args$project$run_cell_cycle
run_cell_cycle=TRUE

if (cell.cycle.string == "FALSE"){
  run_cell_cycle=FALSE
} 
node.type <- config.args$analysis$node_type

part4.suffix <- config.args$analysis$part4$part4_suffix
part5.suffix <- config.args$analysis$part5$part5_suffix

part5.rds.open.file <- config.args$analysis$part5$part5_rds_open_filename
part5.rds.open.file <- sub(".RDS", paste0(part4.suffix, ".RDS"), part5.rds.open.file )
part5.rds.open.path <- file.path(r.obj.loc, part5.rds.open.file)

part5.rds.save.file <- config.args$analysis$part5$part5_rds_save_filename
part5.rds.save.file <- sub(".RDS", paste0(part5.suffix, ".RDS"), part5.rds.save.file )
part5.rds.save.path <- file.path(r.obj.loc, part5.rds.save.file)

part5.file.for.report <- config.args$analysis$part5$part5_report_tables_filename
part5.file.for.report <- sub(".RData", paste0(part5.suffix, ".RData"), part5.file.for.report )
part5.file.report.path <- file.path(r.obj.loc, part5.file.for.report)

ggplot.directory.name <- config.args$analysis$part1$ggplot_dir
ggplot.directory <- file.path(r.obj.loc, ggplot.directory.name)

marker.gene.dir.name <- config.args$analysis$part5$marker_gene_dir
marker.gene.dir.name <- paste0(marker.gene.dir.name, part5.suffix)
marker.gene.dir <- file.path(r.obj.loc, marker.gene.dir.name)

# create ggplot folder if it doesn't exist already
if (!file.exists(ggplot.directory)) {
 dir.create(ggplot.directory, showWarnings = FALSE, recursive = TRUE)
}

if (!file.exists(marker.gene.dir)) {
 dir.create(marker.gene.dir, showWarnings = FALSE, recursive = TRUE)
}
```

```{r define_save_plots, echo = FALSE}

save_png_plot <- function(plot, filename){
  save.path <- file.path(ggplot.directory, paste0("part5", part5.suffix, "_", filename))
  ggsave(save.path, plot)
}

```

### This report was generated on `r formatted.date`

* clustering resolution: `r resolution.selection`
* node.type: `r node.type`

* starting with rds object at `r part5.rds.open.path`
* saving rds object to `r part5.rds.save.path`
* saving file for report at `r part5.file.report.path`
* ggplot figures at `r ggplot.directory`
* Lists of marker genes are at `r marker.gene.dir`

```{r for_testing_only, eval = FALSE, echo = FALSE}

# If you've already run findAllMarkers but want to test other parts
# of this script, then set eval to TRUE here and set the find_all_markers
# code chunk below to eval=FALSE
load(file=part5.rds.save.path)

```


```{r get_num_cores_available, echo = FALSE}
if (node.type == "compute"){
  # If using a compute node, then put this in the submission script:
  # export MC_CORES=${SLURM_NTASKS}
  num.cores.available <- as.numeric(Sys.getenv("SLURM_NTASKS"))
} else {
  library(parallel) 
  num.cores.available <- detectCores() 
}
```

## Load the Seurat object from part 4

```{r load_data, warning=FALSE, error=FALSE, message=FALSE}
experiment.aggregate <- readRDS(file=part5.rds.open.path)
Idents(experiment.aggregate) <- resolution.selection
```


### Choosing a resolution

Table of cluster to sample assignments.
```{r cells_per_res_sample, warning=FALSE, error=FALSE, message=FALSE}
clusters.to.sample.table <- table(Idents(experiment.aggregate),experiment.aggregate$orig.ident) %>% kable(caption = "Number of Cells in each cluster, by sample", jalign = "c") %>% kable_styling()
# table(Idents(experiment.aggregate),experiment.aggregate$orig.ident)
# clusters.to.sample.table <- sapply(grep("res",colnames(experiment.aggregate@meta.data),value = TRUE),
#        function(x) length(unique(experiment.aggregate@meta.data[,x]))) %>% kable(caption = "Number of clusters by resolution", col.names = "Number of clusters", jalign = "c") %>% kable_styling()

clusters.to.sample.table
```

<!-- Plot TSNE coloring by the slot 'ident' (default). -->
<!-- ```{r plot_tsne, warning=FALSE, error=FALSE, message=FALSE} -->
<!-- DimPlot(object = experiment.aggregate, pt.size=0.5, reduction = "tsne", label = T) -->
<!-- ``` -->


### Plot uMap coloring by the slot 'ident' (default).
```{r plot_umap, warning=FALSE, error=FALSE, message=FALSE}
cluster.umap <- DimPlot(object = experiment.aggregate, 
                        pt.size=0.5, 
                        reduction = "umap", 
                        label = T)

cluster.umap

save_png_plot(plot = cluster.umap, filename = "cluster_umap.png")
```

### Plot UMAP  coloring by the slot 'orig.ident' (sample names) with alpha colors turned on. A pretty picture
```{r pretty_pre, warning=FALSE, error=FALSE, message=FALSE}
sample.umap <- DimPlot(object = experiment.aggregate, group.by="orig.ident", pt.size=0.5, reduction = "umap" )

sample.umap
save_png_plot(plot = sample.umap, filename = "sample_umap.png")
```



`r if (run_cell_cycle) {"### uMAP plot by cell cycle"}`
```{r plot_cellcycle, warning=FALSE, error=FALSE, message=FALSE, eval=run_cell_cycle}
cellcycle.umap <- DimPlot(object = experiment.aggregate, pt.size=0.5, group.by = "Phase", reduction = "umap" )

cellcycle.umap
save_png_plot(plot = cellcycle.umap, filename = "cellcycle_umap.png")
```


### Use feature plot to plot our read valued metadata, like nUMI, Feature count, and percent Mito

FeaturePlot can be used to color cells with a 'feature', non categorical data, like number of UMIs
```{r plot_rna, warning=FALSE, error=FALSE, message=FALSE}
nUMI.featureplot <- FeaturePlot(experiment.aggregate, features = c('nCount_RNA'), pt.size=0.5)

nUMI.featureplot
save_png_plot(plot = nUMI.featureplot, filename = "nUMI_featplot.png")
```
and number of genes present
```{r plot_feature, warning=FALSE, error=FALSE, message=FALSE}
nGenes.featureplot <- FeaturePlot(experiment.aggregate, features = c('nFeature_RNA'), pt.size=0.5)

nGenes.featureplot
save_png_plot(plot = nGenes.featureplot, "nGenes_featplot.png")
```

percent mitochondrial
```{r plot_mito, warning=FALSE, error=FALSE, message=FALSE}
pctmito.featureplot <- FeaturePlot(experiment.aggregate, features = c('percent.mito'), pt.size=0.5)

pctmito.featureplot
save_png_plot(plot = pctmito.featureplot, filename = "pctmito_featplot.png")
```


## Identifying Marker Genes

Seurat can help you find markers that define clusters via differential expression.

`FindMarkers` identifies markers for a cluster relative to all other clusters.

`FindAllMarkers` does so for all clusters

`FindAllMarkersNode` defines all markers that split a Node from the cluster tree

### FindAllMarkers can be used to automate the process across all genes.

```{r find_all_markers, warning=FALSE, error=FALSE, message=FALSE, eval=TRUE}
DefaultAssay(experiment.aggregate) <- "RNA"

markers_all <- FindAllMarkers(
    object = experiment.aggregate,
    only.pos = TRUE,
    min.pct = 0.25,
    thresh.use = 0.25
)

markers_all_single <- markers_all[markers_all$gene %in% names(table(markers_all$gene))[table(markers_all$gene) == 1],]

```


pct.1 and pct.2 are the proportion of cells with expression above 0 in ident.1 and ident.2 respectively. p_val is the raw p_value associated with the differntial expression test with adjusted value in p_val_adj. avg_logFC is the average log fold change difference between the two groups.

avg_diff appears to be the difference in log(x = mean(x = exp(x = x) - 1) + 1) between groups.  It doesn’t seem like this should work out to be the signed ratio of pct.1 to pct.2 so I must be missing something.  It doesn’t seem to be related at all to how the p-values are calculated so maybe it doesn’t matter so much, and the sign is probably going to be pretty robust to how expression is measured.


### Feature plots and Violin Plots of Marker Genes
```{r gene_feature, warning=FALSE, error=FALSE, message=FALSE, eval=TRUE, fig.width = 25, fig.height = 20}
feat_plots_top_genes <- function(cluster.num){  
  # cluster.num = 1
  
  saved.fig.width = 25
  saved.fig.height = 20
  
  print(paste0("Cluster number ", cluster.num, ", width: ", saved.fig.width, 
               ", height: ", saved.fig.height))
  curr.markers <- markers_all[markers_all$cluster == cluster.num,] 
  num.markers.to.plot <- 6
  # print(head(curr.markers))

  feat.plot.path <- file.path(ggplot.directory, 
                              paste0("part5", part5.suffix, "_marker_gene_featPlot_cl_", cluster.num, ".png"))
  vln.plot.path <- file.path(ggplot.directory, 
                             paste0("part5", part5.suffix, "_marker_gene_vlnPlot_cl_", cluster.num, ".png"))

  feat.plot.list <- FeaturePlot(
    experiment.aggregate,
    curr.markers$gene[1:num.markers.to.plot],
    cols = c("lightgrey", "blue"),
    ncol = 3,
    pt.size = 2,
    label.size = 30)

  # Fix the axis labels and main title sizes for the feature plots
  for (i in 1:num.markers.to.plot){
    feat.plot.list[[i]] <- feat.plot.list[[i]] + theme(axis.text.x = element_text(angle=45, hjust=1, size=30),
                                                       axis.text.y = element_text(size = 30),
                                                       plot.title = element_text(size=30))
  }

  plot.title <- ggdraw() + draw_label(paste0("Cluster ", cluster.num), fontface = 'bold', size = 30)
  print(cowplot::plot_grid(plot.title, feat.plot.list, ncol = 1, rel_heights = c(0.1, 1)))


  ggsave(feat.plot.path, width=saved.fig.width, height=saved.fig.height)


  vln.plot.list <- VlnPlot(object = experiment.aggregate,
                      features = curr.markers$gene[1:num.markers.to.plot],
                      pt.size = 0.05)

  # Fix the axis labels and main title sizes for the violin plots
  for (i in 1:num.markers.to.plot){
    vln.plot.list[[i]] <- vln.plot.list[[i]] + theme(axis.text.x = element_text(angle=45, hjust=1, size=30),
                                                       axis.text.y = element_text(size = 30),
                                                       plot.title = element_text(size=30))
  }
  print(cowplot::plot_grid(plot.title, vln.plot.list, ncol = 1, rel_heights = c(0.1, 1)))


  ggsave(vln.plot.path, width=saved.fig.width, height=saved.fig.height)


}

# these.markers <- markers_all[markers_all$cluster == 6,]
# VlnPlot(object = experiment.aggregate, features = these.markers$gene[1:6], pt.size = 0.05)
# num.clusters.list = c(0, 1)
num.clusters.list <- unique(markers_all$cluster)
# num.clusters.list
sapply(num.clusters.list, feat_plots_top_genes)
```


<br/>
<br/>

Plot a heatmap of genes by cluster for the top 10 marker genes per cluster
```{r markers_head, warning=FALSE, error=FALSE, message=FALSE, fig.height=10}
DefaultAssay(experiment.aggregate) <- "SCT"
top10 <- markers_all_single %>% group_by(cluster) %>% top_n(10, avg_log2FC)
heatmap.genes.by.cluster <- DoHeatmap(
    object = experiment.aggregate,
    features = top10$gene
)
    
heatmap.genes.by.cluster   

save_png_plot(plot = heatmap.genes.by.cluster, file = "heatmap_genes_by_cluster.png")
```

And last lets save all the objects in our session.
```{r save_rdata, warning=FALSE, error=FALSE, message=FALSE}

save.for.rdata <- c("experiment.aggregate",
                    "markers_all",
                    "markers_all_single",
                    "top10",
                    "num.clusters.list")

save(list=save.for.rdata, file=part5.rds.save.path)
# load(part5.rds.save.path)

to.save.for.report <- c("clusters.to.sample.table",
                        "num.clusters.list")

num.clusters <- length(unique(markers_all$cluster))

for (i in seq(0, num.clusters - 1)){
  table.filename <- file.path(marker.gene.dir, paste0("cluster", i, "_marker_genes_", resolution.selection, ".csv"))
  these.markers <- markers_all[markers_all$cluster == i,]
  write.csv(x = these.markers, file = table.filename, quote = FALSE)
}

# to.save.for.report <- c("marker.gene.plots", "num.clusters.list")
save(list = to.save.for.report, file = part5.file.report.path) 
```

## Session Information
```{r session_info, warning=FALSE, error=FALSE, message=FALSE}
sessionInfo()
```
