---
output:
  html_document:
    pandoc_args: './common.yaml'
params:
  config.args: "./config.yaml"
---

# Part 2: Some QA/QC, filtering, normalization, and assigning variable genes

## Load libraries
```{r libraries, warning=FALSE,error=FALSE,message=FALSE}
library(Seurat)
library(biomaRt)
library(ggplot2)
library(knitr)
library(kableExtra)
library(future)
library(yaml)
```

```{r prepare_inline_variables, echo = FALSE}
# get the datetime and set it to variable
todays.date <- Sys.Date()
formatted.date <- format(todays.date, format="%B %d %Y")

config.args <- read_yaml(params$config.args)

```

```{r user_vals, warning=FALSE,error=FALSE,message=FALSE, echo=FALSE}
r.obj.loc <- config.args$data$`rds-file-path`
organism <- config.args$project$organism

mito.cutoff = config.args$analysis$part2$mito_cutoff
RNA.count.floor = config.args$analysis$part2$RNA_count_floor
RNA.count.ceiling = config.args$analysis$part2$RNA_count_ceiling
feature.count.floor = config.args$analysis$part2$feature_count_floor
feature.count.ceiling = config.args$analysis$part2$feature_count_ceiling
rbc.cutoff = config.args$analysis$part2$rbc_cutoff
ribo.cutoff = config.args$analysis$part2$ribo_cutoff

run_tests <- config.args$analysis$run_tests
node.type <- config.args$analysis$node_type

part1.suffix <- config.args$analysis$part1$part1_suffix
part2.suffix <- config.args$analysis$part2$part2_suffix

# Generally I don't expect part1 to be modified, but I'm sure I'll be proven wrong
# For now don't allow other file suffixes for the part1 output to avoid
# having to generate it for every downstream mod
part2.rds.open.file <- config.args$analysis$part2$part2_rds_open_filename
part2.rds.open.file <- sub(".RDS", paste0(part1.suffix, ".RDS"), part2.rds.open.file )
part2.rds.open.path <- file.path(r.obj.loc, part2.rds.open.file)

#--------------------------------------------------#
# Set up the filenames by taking the normal output
# filename from the config and adding the suffix
#--------------------------------------------------#
part2.rds.save.file <- config.args$analysis$part2$part2_rds_save_filename
part2.rds.save.file <- sub(".RDS", paste0(part2.suffix, ".RDS"), part2.rds.save.file )
part2.rds.save.path <- file.path(r.obj.loc, part2.rds.save.file)

part2.file.for.report <- config.args$analysis$part2$part2_report_tables_filename
part2.file.for.report <-sub(".RData", paste0(part2.suffix, ".RData"), part2.file.for.report)
part2.file.report.path <- file.path(r.obj.loc, part2.file.for.report)

ggplot.directory.name <- config.args$analysis$part1$ggplot_dir
ggplot.directory <- file.path(r.obj.loc, ggplot.directory.name)
```


```{r define_save_plots, echo = FALSE}

save_png_plot <- function(plot, filename){
  save.path <- file.path(ggplot.directory, paste0("part2", part2.suffix, "_", filename))
  ggsave(save.path, plot)
}

```
### This report was generated on `r formatted.date`

* node.type: `r node.type`

* starting with rds object at `r part2.rds.open.path`
* saving rds object to `r part2.rds.save.path`
* saving file for report at `r part2.file.report.path`

* mito cutoff: `r mito.cutoff`
* RNA count floor: `r RNA.count.floor`
* RNA count ceiling: `r RNA.count.ceiling`
* Feature count floor: `r feature.count.floor`
* Feature count ceiling: `r feature.count.ceiling`
* RBC cutoff: `r rbc.cutoff`
* Ribosomal cutoff: `r ribo.cutoff`



```{r get_num_cores_available, echo = FALSE}
if (node.type == "compute"){
  # If using a compute node, then put this in the submission script:
  # export MC_CORES=${SLURM_NTASKS}
  num.cores.available <- as.numeric(Sys.getenv("SLURM_NTASKS"))
} else {
  library(parallel)
  num.cores.available <- detectCores() 
}
```

## Load the Seurat object from part 1

```{r load_data, warning=FALSE,error=FALSE,message=FALSE}
experiment.aggregate <- readRDS(file=part2.rds.open.path)
set.seed(12345)
```


<br/>

## Before Filtering

Ridgeplots before applying filters

```{r generate_ridgeplot_pre, warning=FALSE,error=FALSE,message=FALSE}
ridgeplot.prefiltering <- RidgePlot(experiment.aggregate, features=c("nFeature_RNA","nCount_RNA", "percent.mito", "percent.rbc", "percent.ribo"), ncol = 2)
# RidgePlot(experiment.aggregate, features=c("nFeature_RNA","nCount_RNA", "percent.mito"), ncol = 2)
```

<br/>

## Proposed filtering cutoffs

Gene Plot, scatter plot of gene expression across cells, (colored by sample), drawing horizontal an verticale lines at proposed filtering cutoffs.

```{r relationships, warning=FALSE,error=FALSE,message=FALSE}
# Set these cutoffs here and use them in the plots AND in the filtering

nCount.mito.scatter <- FeatureScatter(experiment.aggregate, "nCount_RNA", "percent.mito") + 
  geom_vline(xintercept = c(RNA.count.floor,RNA.count.ceiling)) + 
  geom_hline(yintercept = mito.cutoff)

nFeature.mito.scatter <- FeatureScatter(experiment.aggregate, "nFeature_RNA", "percent.mito") + 
  geom_vline(xintercept = c(feature.count.floor, feature.count.ceiling)) + 
  geom_hline(yintercept = mito.cutoff)

nCount.nFeature.scatter <- FeatureScatter(experiment.aggregate, "nCount_RNA", "nFeature_RNA")  + 
  geom_vline(xintercept = c(RNA.count.floor,RNA.count.ceiling)) + 
  geom_hline(yintercept = c(feature.count.floor, feature.count.ceiling))

ribo.rbc.scatter <- FeatureScatter(experiment.aggregate, "percent.ribo", "percent.rbc")+ 
  geom_vline(xintercept = ribo.cutoff) + 
  geom_hline(yintercept = rbc.cutoff)

ribo.mito.scatter <- FeatureScatter(experiment.aggregate, "percent.ribo", "percent.mito")+ 
  geom_vline(xintercept = ribo.cutoff) + 
  geom_hline(yintercept = mito.cutoff)

mito.rbc.scatter <- FeatureScatter(experiment.aggregate, "percent.mito", "percent.rbc")+ 
  geom_vline(xintercept = mito.cutoff) + 
  geom_hline(yintercept = rbc.cutoff)

nFeature.rbc.scatter <- FeatureScatter(experiment.aggregate, "nFeature_RNA", "percent.rbc")+ 
  geom_vline(xintercept = c(feature.count.floor, feature.count.ceiling)) + 
  geom_hline(yintercept = rbc.cutoff)

nCount.mito.scatter 
nFeature.mito.scatter 
nCount.nFeature.scatter 
ribo.rbc.scatter 
ribo.mito.scatter 
mito.rbc.scatter 
nFeature.rbc.scatter 

save_png_plot(plot = nCount.mito.scatter, filename = "nCount_mito.png")
save_png_plot(plot = nFeature.mito.scatter, filename = "nFeat_mito.png")
save_png_plot(plot = nCount.nFeature.scatter, filename = "nCount_nFeat.png")
save_png_plot(plot = ribo.rbc.scatter, filename = "ribo_rbc.png")
save_png_plot(plot = ribo.mito.scatter, filename = "ribo_mito.png")
save_png_plot(plot = mito.rbc.scatter, filename = "mito_rbc.png")
save_png_plot(plot = nFeature.rbc.scatter, filename = "nFeat_rbc.png")
```

## Cell filtering stats

We use the cutoffs above to filter out cells. 

```{r apply_filtering, warning=FALSE,error=FALSE,message=FALSE}
before.table <- table(experiment.aggregate$orig.ident)

experiment.aggregate <- subset(experiment.aggregate, percent.mito <= mito.cutoff)
experiment.aggregate <- subset(experiment.aggregate, nCount_RNA >= RNA.count.floor & nCount_RNA <= RNA.count.ceiling)
experiment.aggregate <- subset(experiment.aggregate, nFeature_RNA >= feature.count.floor & nFeature_RNA <= feature.count.ceiling)
experiment.aggregate <- subset(experiment.aggregate, percent.rbc >= rbc.cutoff)
experiment.aggregate <- subset(experiment.aggregate, percent.ribo <= ribo.cutoff)

after.table <- table(experiment.aggregate$orig.ident)

combo.table <- rbind(before.table,
                     after.table)

row.names(combo.table) <- c("Num cells before filtering", "Num cells after filtering")

pretty.combo.table <- combo.table %>%
  kable(caption = "Cells present before and after filtering")%>%
  pack_rows("Filtering results", 1, 2, label_row_css = "background-color: #666; color: #fff;") %>%
  kable_styling("striped")

pretty.combo.table
```

<br/>

## Filtering comparison 

Ridgeplots before filtering:
```{r ridgeplot_pre, warning=FALSE,error=FALSE,message=FALSE}
ridgeplot.prefiltering
save_png_plot(plot = ridgeplot.prefiltering, 
              filename = "ridgeplot_prefiltering.png")
```


Ridgeplots after filtering
```{r ridgeplot_post, warning=FALSE,error=FALSE,message=FALSE}
ridgeplot.postfiltering <- RidgePlot(experiment.aggregate, features=c("nFeature_RNA","nCount_RNA", "percent.mito", "percent.rbc", "percent.ribo"), ncol = 2)
ridgeplot.postfiltering

save_png_plot(plot = ridgeplot.postfiltering, 
              filename = "ridgeplot_postfiltering.png")
```

<br/>
## Finally, lets save the filtered and normalized data
```{r save_r, warning=FALSE,error=FALSE,message=FALSE}
saveRDS(experiment.aggregate, file=part2.rds.save.path)

to.save.for.report <- c("mito.cutoff", 
                        "RNA.count.floor", 
                        "RNA.count.ceiling", 
                        "feature.count.floor", 
                        "feature.count.ceiling", 
                        "rbc.cutoff", 
                        "ribo.cutoff",
												"pretty.combo.table"
                        )

save(list = to.save.for.report, file=part2.file.report.path)
```

## Session Information
```{r session_info, warning=FALSE,error=FALSE, message=FALSE}
sessionInfo()
```

